Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-02 linear_iso.tex
revised analysis part

\section{Analytical discussion of vertically isothermal disks}\label{analytical}
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$), before considering the effect of thermal relaxation.

@@

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

@@

We will consider sound-speed profiles with $q<0$ in the global 
disk. Then $\Omega^2(z)$ decreases away from the midplane,
and may reach small values for sufficiently large $|z|$, which
invalidates the low-frequency approximation. In principle, then, the
above approximations implicitly limits us to consider thin disks ($\epsilon
\ll 1$) with vertical domain sizes not too large, so that the rotation
is nearly-Keplerian.

\subsection{Isothermal perturbations}\label{iso_discuss}
Isothermal perturbations corresponds to $\gamma=1$ for any thermal
relaxation timescale $\beta$, or for any adiabatc index $\gamma$ when
$\beta=0$. In this case Eq. \ref{ode_Q} implies 
\begin{align}
  Q = W, 
\end{align}
 and so 
 \begin{align}
   \ii\hat{\sigma}c_s\delta v_z = W^\prime
   % \quad   \text{(isothermal perturbations)},
 \end{align}
from Eq. \ref{lin_vz}. For isothermal perturbations 
the parameter $\chi=1$ so it is easiest to work with
$W$ by integrating Eq. \ref{vertiso_gov_nondim}. For later discussion
it is convenient to write the result as

>>>>

\section{Analytical discussion of vertically isothermal disks}\label{analytical}
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
first examine the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $\beta$, or any value of $\gamma$ with
$\beta = 0$) before considering the effect of thermal relaxation. In the
final section we consider adiabatic disks ($\gamma>1$ and
$\beta\to\infty$) to demonstrate an important limitation of the 
local model.

@@

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  % &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
  %     \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
  &+ \left[\frac{\sigma^2}{c_s^2}\left(\frac{k_x^2c_s^2}{\Omega_k^2} +
      \chi\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

@@

% We will consider sound-speed profiles with $q<0$ in the global 
% disk. Then $\Omega^2(z)$ decreases away from the midplane,
% and may reach small values for sufficiently large $|z|$, which
% invalidates the low-frequency approximation. In principle, then, the
% above approximations implicitly limits us to consider thin disks ($\epsilon
% \ll 1$) with vertical domain sizes not too large, so that the rotation
% is nearly-Keplerian.

Note that we can generalize Eq. \ref{vertiso_gov_nondim} to other
(non-Keplerian) rotation profiles by redfining $\khat$
 as $k_xc_s/\kappa$, $\hat{\sigma}$ as $\sigma/\kappa$, and 
 $\epsilon$ as $c_s/r\kappa$.

% Consider rotation profiles with radial dependence
% $\propto r^{-m}$. Then the same dimensionless governing equation 
% (Eq. \ref{vertiso_gov_nondim}) holds, provided we redfine $\khat$
% as $k_xc_s/\kappa$, $\hat{\sigma}$ as $\sigma/\kappa$, and reset
% $\epsilon$ to $\epsilon/sqrt{4-2m}$.

\subsection{Isothermal perturbations}\label{iso_discuss}
Isothermal perturbations corresponds to $\gamma=1$ for any thermal
relaxation timescale $\beta$, or for any adiabatc index $\gamma$ when
$\beta=0$. In either case $\chi=1$ and Eq. \ref{ode_Q} implies 
\begin{align}
  Q = W, 
\end{align}
 and so 
 \begin{align}
   \ii\hat{\sigma}c_s\delta v_z = W^\prime
   % \quad   \text{(isothermal perturbations)},
 \end{align}
from Eq. \ref{lin_vz}. For isothermal perturbations 
the parameter $\chi=1$ so it is easiest to work with
$W$ by integrating Eq. \ref{vertiso_gov_nondim}. For later discussion
it is convenient to write the result as

### 2015-02-02 setup.tex
revised analysis part

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.  
% From $\epsilon$ we can define a
% characteristic scale-height  
% \begin{align}
%   \Hiso \equiv \epsilon r.% \quad\quad(\Gamma \equiv 1),   
% \end{align} 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $\gamma$ is the constant adiabatic index. We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

% {\bf need to mention: generally consider sub-keplerian disks -> caveat is that it's entirely possble for angular  
% velocity to reach zero at finite height. give example later for isothermal disk. this limits domain size, but this critical height is usually
% very large for thin disks}

@@

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p R} \right)&>0.  
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we evaluate the second criterion
for the above disk models in $z>0$ where $\p_zP<0$, without loss of
generality,   
\begin{align}
  \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
    -\left(\gamma-\Gamma\right)\frac{\p\ln{P}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, to have instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

@@

In this shearing box the equilibrium density, pressure and
sound-speed are assumed to depend only on $z$, by evaluating these
quantities at $r=r_0$ in the global disk. The background rotation is
approximated by a linear shear in $x$, but left unaltered in
$z$. However, the vertical shear $\p_z\Omega$ is also assumed to only
depend on $z$.

@@

\subsection{Limitations}
Before using the VGSB to study the VSI, it is very important to be
aware of several limitations of the box model. The VGSB momentum  
equation have the same form as that for the standard shearing box 
\citep{goldreich65}, except for the appearence of the term 
$v_zr_0d\Omega_0/dz$. This term appears in the global stability
problem and is the only term associated with vertical shear that is 
ported from the global disk into the local model.

%horizontal structure not ported

Note that, within the VGSB framework, the vertical shear 
$d\Omega_0/dz$ is a free parameter (as are $\Omega_0$ and $S_0$). 
This is different to the global problem because in that case, the
vertical shear is a result of the disk radial structure, which is
ignored in the VGSB equilibria. 
%because there is only one
%equilibrium condition, which determines the vertical density/pressure
%structure. 
In effect, we are studying dynamics in the standard shearing box
subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$. This
force is inspired by the vertical shear in the global disk, and its
functional form is taken from global equilibrium. 
%but should be considered as a model parameter

@@

In Appendix we make the connection between the linear problem in the
VGSB to that in the global disk model.

We can eliminate the perturbed velocities between the linearized
equations to obtain a pair or ordinary differential equations
for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
    k_x r}{D}\frac{d\Omega^2}{dz} -
  \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
  \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
&- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
&\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
\frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
&=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
\left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
 % It is convenient to express $\delta v_x$ as a function of $W$ and
 % $\delta v_z$,
 % \begin{align}
 %   \delta v_x = -\frac{1}{D}\left(\sigma k_x W +
 %     r\frac{d\Omega^2}{dz}\delta v_z\right),
 % \end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2.
\end{align}
% Eq. \ref{ode_w}---\ref{ode_Q} are convenient for numerical work, 
% but alternative forms are more suitable for analytical discussion, and 
% will be derived later.

%These boundary conditions are also numerically
%convenient because they can be recast to contain the eigenfrequency
%$\sigma$. This means that the the full set of ODEs, including boundary
%conditions can be written in the form $L(\bm{U}) = \sigma R (\bm{U})$

>>>>

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.

The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $\gamma$ is the constant adiabatic index. We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

@@

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p R} \right)&>0.  
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we evaluate the second criterion
for the above disk models in $z>0$ where $\p_zP<0$, without loss of
generality,   
\begin{align}\label{solberg2}
  \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
    -\left(\gamma-\Gamma\right)\frac{\p\ln{P}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, to have instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

@@

In this shearing box the equilibrium density, pressure and
sound-speed are assumed to depend only on $z$, by evaluating these
quantities at $r=r_0$ in the global disk. The background rotation
$\Omega$ is approximated by a linear shear in $x$, but left unaltered
in $z$. However, the vertical shear $\p_z\Omega$ is also assumed to
only depend on $z$.

@@

We can eliminate the perturbed velocities between the linearized
equations to obtain a pair or ordinary differential equations
for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
    k_x r}{D}\frac{d\Omega^2}{dz} -
  \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
  \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
&- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
&\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
\frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
&=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
\left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2.
\end{align}

\subsection{Limitations}
Before using the above equations to study the VSI, it is very important to be
aware of several limitations of the VGSB model. The VGSB momentum  
equations have the same form as that for the standard shearing box 
\citep{goldreich65}, except for the appearence of the term 
$v_zr_0d\Omega_0/dz$. This term appears in the global stability
problem and is the only term associated with vertical shear that is 
ported from the global disk into the local model.

Note that, within the VGSB framework, the vertical shear 
$d\Omega_0/dz$ is a free parameter (as are $\Omega_0$ and $S_0$). 
This is different to the global problem because in that case, the
vertical shear is a result of the disk radial structure, which is
ignored in the VGSB equilibria. 
In effect, we are studying dynamics in the standard shearing box
subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$. This
force is inspired by the vertical shear in the global disk, and its
functional form is taken from global equilibrium.

@@

In Appendix \ref{global_corr} we make the connection between the
linear problem in the VGSB to that in the global disk model. We
emphasize that not all calculations in the local model can be 
consistently interpreted in the global context. However, for
perturbations with small radial lengthscales (large $|k_x|$) and fast
thermal relaxation ($\beta\ll1$), which characterizes the VSI, the
VGSB framework is adequate.
